---
title: "lab_12"
author: "Derek Willis"
date: "2025-01-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

-   tidytext and our usual libraries

## Load libraries and establish settings

**Task** Create a codeblock and load appropriate packages and settings for this lab.

```{r}
library(tidyverse)
library(tidytext)
library(janitor)
library(lubridate)
library(rvest)
```

## Questions

**Q1.** You've been assigned to report a story about the leading reasons that Maryland attorneys get sanctioned by the state for misconduct. The state [publishes lists of sanctions](https://www.courts.state.md.us/attygrievance/sanctions) that contain a short text description about the situation. Load the CSV file in the data folder containing records from fiscal year 2011 onwards. Make a list of unique words from the text column, then following the example in the pre_lab, remove common "stop words" from that list and create a list of the top 10 words containing the percentage of occurrences each word represents. What's the leading word in that answer and, broadly, what do you think the top 10 words describe?

**A1. The leading word is "failing" at just over 4.5%, and the top 10 words overall seem to have a theme of people failing to uphold their end of a deal in a legal matter (failing, failed, client, respondent, etc)**

```{r}

md_attorney_sanctions <- read_csv("data/md_attorney_sanctions.csv")

unique_words <- md_attorney_sanctions |> select(text) |>
  unnest_tokens(word, text)

unique_words |>
  anti_join(stop_words) |>
  group_by(word) |>
  tally(sort=TRUE) |>
  mutate(percent = (n/sum(n))*100) |>
  top_n(10)
```

**Q2.** Let's move beyond single words to phrases. Make a list of the top 10 three-word phrases, called trigrams, based on the example from the pre_lab (you'll need to modify the example code to do this). What's the top trigram and how often does it appear? What does that phrase mean in legal terms?

**A2.** The top trigram is "attorney trust account" at about 2.4%. In a legal sense, this is a special bank account to separate lawyer's funds from their clients. (I think that's what it means, Google was a bit unclear).

```{r}

trigrams <- md_attorney_sanctions |>
  unnest_tokens(trigram, text, token = "ngrams", n = 3) |>
  separate(trigram, c("word1", "word2", "word3"), sep = " ") |>
  filter(!word1 %in% stop_words$word) |>
  filter(!word2 %in% stop_words$word) |>
  mutate(trigram = paste(word1, word2, word3, sep=" ")) |>
  group_by(trigram) |>
  tally(sort=TRUE) |>
  mutate(percent = (n/sum(n))*100) |>
  top_n(10)
```

**Q3.** Let's drop back down to more traditional text analysis - take the top trigram from Q2 and write code to see how many times it occurs in the text column in each fiscal year. What do you think the answer produced by your code suggests? What else could you do to try and clarify the most important reasons attorneys get sanctioned?

**A3. The highest amount was in 2021, but there were no occurrences in 2012, 2013 or 2014 (but there were a few in 2011). I guess it means that lawyers were pretty responsible for a few years in terms of keeping their accounts in order then really started slacking off again. The way I could find out the most important reasons attorneys get sanctioned is go back to the ngram list from question two and see how often each one happens for each fiscal year, figure out what reasons have grown or shrunk in their occurrence rate.**

```{r}

trigram_occur <- md_attorney_sanctions |>
  mutate(has_trigram = str_detect(text, "attorney trust account|Attorney trust account|Attorney Trust Account|ATTORNEY TRUST ACCOUNT")) |>
  group_by(fiscal_year) |>
  summarise(count = sum(has_trigram)) |>
  arrange(desc(count))
```
