---
title: "lab_03"
author: "Derek Willis"
date: "2025-01-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## TASK: Load libraries

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse
library(tidyverse)
# Load janitor
library(janitor)
```

## TASK: Load data and clean up the column names

Load the two datasets we worked with in the pre_lab that are in the data folder: the Maryland payments to vendors and Maryland grants and loans AND, using janitor, change the column names so that they are lower-cased and any spaces are replaced with underscores.

```{r}
# Load in the Maryland state agency payments data, making sure that it correctly sets the datatypes using all of the rows

payments <- read_csv("data/State_of_Maryland_Payments_Data__FY2008_to_FY2024.csv") |> 
  clean_names()

# Load in the Maryland grants and loan data

md_grants_loans <- read_csv("data/State_of_Maryland_Grant_and_Loan_Data__FY2009_to_FY2022.csv")|> 
  clean_names()
```

## Answer questions

Q1: You want to know how much money state agencies have paid to the Prince George's County School system, but to do that you'll need to find all of the variations of how the school system might be listed in the payments data. Look at & search the records in the dataframe to start your search, starting with a broad search and then narrowing it down. Then write code to find rows where the vendor name contains some version of the county's school system.

HINT: you should use filter() just once, and inside your filter you want to look for the variations one at a time. Think "this version OR that version."

When you have found as many as possible, save the result to a new dataframe called `pg_schools`. How many records did you find?

A1: 98 records, however that could include some false positives, plus I also included "education" so get board of education in the dataset

```{r}
pg_schools <- payments |>
  filter(
    str_detect(vendor_name, "PG|PRINCE GEORGE") &
    str_detect(vendor_name, "SCHOOL|EDUCATION")
  )
```

Q2: Using that `pg_schools` dataframe, write code to calculate the total amount of payments to the Prince George's County School system for each fiscal year. Which fiscal year had the largest total amount? Are there any aspects of the results that stand out to you?

A2: The 2024 fiscal year has the highest total amount. The thing that stands out the most is that 2025 is the lowest amount, about $770 million less than the prior year. The biggest issue here is I don't know when this data was collected, which could mean that 2025 is under-reported in this dataset.

```{r}
pg_per_year <- pg_schools |> group_by(fiscal_year) |> 
  summarise(total_amount = sum(amount)) |> 
  arrange(desc(total_amount))
```

Q3. You're interested in examining the vendors paid for capital projects at the University of Maryland. Search the payments data to find those payments, then write code to produce a list of each vendor and the total amount it received.

Look carefully at that list - what cleaning will it require to ensure that you have only one row per vendor and zip code? Provide at least three examples of vendor names that are similar and should be combined into one. Describe in words how you might clean that data so that any nearly identical vendors would look identical - what would you change and how?

A3. The first one that jumps out is the second- and third-ranked agencies are "WHITING TURNER CONTRACTING CO" and "WHITING TURNER CONTRACTING COMPANY." Cleaning will need to be done to make sure R knows those are the same, you could use something like the casewhen() function to try and fix that issue and let R know that CO is the same as COMPANY. Similar issues that could require the casewhen() are "AND," "+" and "&," or commas located within agency names.

```{r}
capital_proj_umd <- payments |> 
  filter(str_detect(agency_name, "CAPITAL PROJECTS- UNIVERSITY OF MARYLAND")) |>
  group_by(vendor_name) |> 
  summarise(total_amount = sum(amount)) |> 
  arrange(desc(total_amount))
```

Q4: You want to understand how much state grant money went to stem cell research projects each fiscal year. Search the data to find grants that fit that description. Once you find which column(s) contain the information you want, your code should find just the grants for stem cell research in a single filter. Then calculate the total amount of money for each fiscal year. Write a sentence summarizing how stem cell research funding has changed over time.

A4: The most spending on stem cell research came in 2009, then it slowly dropped over time before bottoming out in 2021 and shooting back up from 2022 to 2023. Potential errors here could be people who misspelled "stem cell" or forgot to list their description in the proper column.

```{r}
stem_cell <- md_grants_loans |>
  filter(str_detect(description, "STEM CELL|Stem Cell|stem cell")) |>
  group_by(fiscal_year) |> 
  summarise(total_amount = sum(amount)) |> 
  arrange(desc(total_amount))
```

Q5: Using the Maryland grants and loan data, find records where the zip code is the one where the University of Maryland-College Park is located in and produce a list of grantees from that zip code, along with the total amount they received. Write a paragraph explaining how many grantees are there, what you would need to do to clean the results so that you had one record per grantee and which grantees are most interesting to you?

A5: There are 51 grantees in the initial list. That said, it requires some cleaning because there are some grantees that are listed in both caps/lowercase and ALL CAPS, making R see them as different grantees. Other errors include what I saw in task three, where some orgs abreviate differently (Co vs Company vs Co. or nothing at all, sometimes they also omit LLC or INC). This is another situation where casewhen() would be helpful, but to fix the issue with the letters you can use the janitor tool str_to_upper and make everything in the grantee column uppercase to fix that problem.

Potential errors here would be people who used the wrong zip code or people who added the "plus four," my code would not be able to find it in that case

```{r}
college_park_zip <- md_grants_loans |>
  filter(str_detect(zip_code, "20740|20741|20742")) |>
  group_by(grantee) |> 
  summarise(total_amount = sum(amount)) |> 
  arrange(desc(total_amount))
```
