---
title: "final_project"
output: html_document
---

```{r}
library(tidyverse)
library(janitor)

wbb_teams <- read_csv("womens-college-basketball-main/teams.csv") |> clean_names()

wbb_rosters23 <- read_csv("womens-college-basketball-main/wbb_rosters_2022_23.csv") |> mutate(player_id = as.character(player_id))
wbb_rosters24 <- read_csv("womens-college-basketball-main/wbb_rosters_2023_24.csv") |> mutate(player_id = as.character(player_id))
wbb_rosters25 <- read_csv("womens-college-basketball-main/wbb_rosters_2024_25.csv") |> mutate(player_id = as.character(player_id))
wbb_rosters26 <- read_csv("womens-college-basketball-main/wbb_rosters_2025_26.csv")

combined_rosters <- bind_rows(list(wbb_rosters23, wbb_rosters24, wbb_rosters25, wbb_rosters26)) # <- This code line puts all the roster info into one dataframe

non_us <- read_csv("womens-college-basketball-main/non_us.csv")|> clean_names()

coach_history <- read_csv("womens-college-basketball-main/coaches/coaches_with_history.csv") |> clean_names()
```
Data exploration: Once all the roster data is combined, we have over 41000 data points across 31 different variables. While some of those 31 are pretty useless, (NCAA ID, Player ID, etc), some things we didn't expect it to include like high school and hometowns were in the dataset.

The other dataframes we got were the teams, which doesn't have a ton of relevant information at first glance, but does show us that there are over 1100 schools that make up this dataset, a non-us chart that shows the amount of players from every county and the percent change from 2023 to 2024, then lastly there's also a coaching history dataframe. This one has almost 13000 observations, stating the person's title on the staff, their school and tenure.

As far as the limitations, this data only goes back four years. Not a huge sample of time to look at, but there is a LOT of data for the four years we have access to in this project. Because of that timeframe, we can't really answer the NIL questions we posed last week and we will need to adjust our time frames for the change over time questions.

Q1: Do P5 schools get more international players than mid-majors?

By total count, mid/low major schools get more international players. But, by proportion, P5 schools atract much more international players

```{r}
combined_rosters_international <- combined_rosters |>
  mutate(
    conference_type = case_when(
      str_detect(conference, "ACC") ~ "Power5",
      str_detect(conference, "Big East") ~ "Power5",
      str_detect(conference, "Big Ten") ~ "Power5",
      str_detect(conference, "Big 12") ~ "Power5",
      str_detect(conference, "Pac-12") ~ "Power5",
      str_detect(conference, "SEC") ~ "Power5",
      .default = "Mid/Low"
    ),
    international = case_when(
      str_detect(country_clean, "USA") ~ "No",
      .default = "Yes"
    )
  )

combined_rosters_international |>
  group_by(conference_type, international) |>
  summarise(count = n())

combined_rosters_international |>
  group_by(conference_type) |>
  summarise(
    total_players = n(),
    international_players = sum(international == "Yes"),
    percent_international = international_players / total_players * 100
  )
```
Q2: Head coaching turnover?

Overall, more schools opted to switch their head coach than hold on, but it was very close. There were 833 instances of a head coaching change in the data compared to 717 data points where the coach stayed through the entire timeframe. 
```{r}
coaching_turnover <- coach_history |>
  filter(position_title_standardized == "Head Coach") |>
  filter(category == "College") |>
  mutate(coach_change = !is.na(end_year))

coaching_turnover |>
  group_by(coach_change) |>
  summarise(total = n())
```
Q3: What is the geographic breakdown for the most popular state? Does the income play a role in that geography?

Pennsylvania is the most popular state, with the hotbeds of talent concentrated around Pittsburgh and Philadelphia. To be precise, the most popular city is Houston, just edging out Pittsburgh.

Income did not seem to have a major relationship with where players were from. The hotbed areas did not seem to stand out in any noticable pattern when compared to median income.
```{r}
# First find the most popular state, then build a map of the state

library(tigris)
library(tidycensus)

popular_state <- combined_rosters |>
  group_by(state_clean) |>
  summarise(total = n()) |>
  arrange(desc(total))
popular_state

pa_cities <- places(state = "PA", cb = TRUE) |> clean_names() |>
  left_join(combined_rosters, join_by(name == hometown)) |>
  group_by(name) |>
  summarise(total = n()) |>
  arrange(desc(total))

ggplot() +
  geom_sf(data = pa_cities, aes(fill = total), color = NA) +
  scale_fill_viridis_c(option = "magma") +
  theme_minimal()

pa_city_income <- get_acs(
  geography = "place",
  variables = "B19013_001",
  state = "PA",
  year = 2023,
  survey = "acs5"
)

pa_city_income <- pa_city_income |>
  mutate(
    city = str_remove(
      NAME,
      " (city|borough|town|township|CDP|village),.*$"
    )
  )

pa_cities_income_map <- pa_cities |>
  left_join(pa_city_income, join_by(name == city))

ggplot() + geom_sf(data = pa_cities_income_map, aes(fill = estimate), color = NA) + scale_fill_viridis_c(option = "magma") + theme_minimal()
```
Q4: Was there a change in the amount of four-year seniors pre- and post-NIL?


```{r}
wbb_rosters21 <- read_csv("womens-college-basketball-main/wbb_rosters_2020_21.csv")

pre_nil_seniors <- wbb_rosters21 |>
  filter(str_detect(year, "Senior|Graduate|Graduate Student|Fifth Year|Redshirt Senior|Sr.|R-Sr.|Gr.")) |>
  mutate(
    fouryearsenior = case_when(
      is.na(previous_school_clean) ~ "Yes",
      .default = "No"
    ))

pre_nil_seniors |>
  group_by(fouryearsenior) |>
  summarise(fouryearseniors_pre = n())

post_nil_seniors <- wbb_rosters26 |>
  filter(str_detect(year, "Senior|Graduate|Graduate Student|Fifth Year|Redshirt Senior|Sr.|R-Sr.|Gr.")) |>
    mutate(
    fouryearsenior = case_when(
      is.na(previous_school_clean) ~ "Yes",
      .default = "No"
    ))

post_nil_seniors |>
  group_by(fouryearsenior) |>
  summarise(fouryearseniors_post = n())
```

Q5: How many players stay in their home state?

Over 32000 of the players in the data left their home state to play basketball compared to just under 25,000 who stayed home.
```{r}
home_state_stay <- combined_rosters |>
  left_join(wbb_teams, join_by(team)) |>
  mutate(stay_home = replace_na(state_clean == team_state.y, FALSE))

home_state_stay |>
  group_by(stay_home) |>
  summarise(total = n()) |>
  arrange(desc(total))
```

