---
title: "week7_recap"
author: "Daniel Trielli"
date: "2023-10-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Turn off scientific notation
options(scipen=999)
```

## Loading the packages

Run the codeblock below to load the packages we will need for this recap

```{r}
library(tidyverse)
library(lubridate)
library(janitor)
```

## Load Data

Run the codeblock below to load the data.

```{r}
earthquakes <- read_csv('https://earthquake.usgs.gov/earthquakes/feed/v1.0/summary/all_month.csv')

#Setting time column as datetime
earthquakes <- earthquakes |> mutate(time = as_datetime(time))
```

#### Answer the questions below

Most questions have a code block and a space for an answer below. Write the code you think is necessary and, in the answer space, write out what you did and what was the result.

------------------------------------------------------------------------

#### **Q1** Do you notice anything different with this read_csv()? Particularly in the file name?

Yes, the csv comes from a URL instead of a pre-downloaded file, meaning that it can be updated real-time

------------------------------------------------------------------------

#### **Q2** How many records there are there in this dataset? What do they mean and what useful information we can gather from it, looking at the columns?

As of this moment, there was 10,309 observations in the dataset. This means that there are 10,309 recorded earthquakes from the last month. Some of the interesting information that's useful for identifying newsworthy earthquakes is magnitude, latitude, longitude and depth.

------------------------------------------------------------------------

#### **Q3** How do I reorganize this list to see the ones that are the deepest first? What is the depth that shows up for the deepest one, and its magnitude?

```{r}
earthquakes |> arrange(desc(depth))
```

The deepest one that pops up is 667.197 kilometers. That earthquake had a magnitude of 4.30

------------------------------------------------------------------------

#### **Q4** I just want to see the earthquakes with a magnitude larger than 6. How do I do that? And how many are there that fit this criteria?

```{r}
big_quakes <- earthquakes |>
  filter(mag > 6)
```

There are nine earthquakes that fit that criteria

------------------------------------------------------------------------

#### **Q5** What about if I want to see earthquakes that have both a magnitude larger than 6 and a depth smaller than 20? How many are there in the data set that fit [both]{.underline} these criteria?

```{r}
big_quakes_small_depth <- earthquakes |>
  filter(mag > 6 & depth < 20)
```

There are four earthquakes that fit both criteria

------------------------------------------------------------------------

#### **Q6** What about if I want to see earthquakes that either have a magnitude larger than 6 and a depth smaller than 20? How many are there in the data set that fit [either]{.underline} these criteria?

```{r}
big_quakes_or_small_depth <- earthquakes |>
  filter(mag > 6 | depth < 20)
```

There are 8,251 that fit this new, looser description

------------------------------------------------------------------------

#### **Q7** I'm interested in finding earthquakes that took place in Alaska. Which column should I look at? How do I use it to find all the earthquakes in Alaska? How many earthquakes took place there?

```{r}
alaska_quakes <- earthquakes |>
  filter(str_detect(place, "Alaska|ALASKA|alaska|AK"))
```

You want to use the "place" column. There are 3,330 recorded earthquakes for Alaska

------------------------------------------------------------------------

#### **Q8** I notice that there is a column called 'type', that seems to have different kinds of tremors. What is the best way to find what are all the possible types of tremors, and counting how many of each there are in this data set? What are the first two most common types of tremors in this data set?

```{r}
types <- earthquakes |>
  group_by(type) |>
  summarise(total = n()) |>
  arrange(desc(total))
```

The two most common types in the dataset are earthquake and quarry blast
------------------------------------------------------------------------

#### **Q9** How do I calculate the average depth of each type of tremor? What is the average depth of the earthquake type in this data set? Is there anything that seems unusual or surprising in this finding?

```{r}
depth_per_type <- earthquakes |>
  group_by(type) |>
  summarise(mean_depth = mean(depth)) |>
  arrange(desc(mean_depth))
```

The average depth of the "earthquake" quakes is about 21.87 kilometers. Something interesting is that two categories, (quarry blast and explosion), have negative average depths.

------------------------------------------------------------------------

#### **Q10** I'm interested, in the future, to see the hours in which earthquakes happen. How can I extract the hour from the time column?

```{r}
hourly_quakes <- earthquakes |>
  mutate(hour = hour(time))

hourly_quakes |> group_by(hour) |>
  summarise(total = n()) |>
  arrange(desc(total))
```

Extract is using a mutate function, also make sure lubridate package is running

------------------------------------------------------------------------

#### **Q11** I want to make a note of all the records in this data set that I consider serious. For my purposes, I'm thinking that all tremors that have a magnitude that is larger than 3 are serious. How do I automatically create a new column making that distinction?

```{r}
bigger_than_three <- earthquakes |>
  mutate(
    serious_quake = case_when(
      mag > 3 ~ "SERIOUS",
      .default = "NOT SERIOUS"))
```

Use this case_when function to find your serious/not serious quakes

------------------------------------------------------------------------

#### **Q12** I have no idea how earthquakes work and I'm interested in seeing if there is a particular time of day in which serious earthquakes happen. How can I see that condensed in a table with all the hours in a day and all the serious earthquakes in each hour? What is the hour with fewer serious earthquakes and the one with the most serious earthquakes?

```{r}
final_question <- earthquakes |>
  mutate(
    serious_quake = case_when(
      mag > 3 ~ "SERIOUS",
      .default = "NOT SERIOUS")) |>
  mutate(hour = hour(time))

final_question |> filter(serious_quake == "SERIOUS") |>
  group_by(hour) |>
  summarise(total = n()) |>
  arrange(desc(total))
```

The hour with the most serious quakes is 17 with 67, while the hour with the fewest serious quakes is 2 with 38

------------------------------------------------------------------------

#### EXTRA Q What other questions are you interested in and how would you ask them in R?
How many of each magType is there?
```{r}
earthquakes |>
  group_by(magType) |>
  summarise(total = n()) |>
  arrange(desc(total))

#There are nine different mag types, the most common is ml with 6966
```
What is the largest depth error is the dataset?
```{r}
earthquakes |> arrange(desc(depthError))

# The largest depth error seen is 125.6 kilometers.
```
How many earthquakes received an automatic review?
```{r}
earthquakes |>
  group_by(status) |>
  summarise(total = n()) |>
  arrange(desc(total))

# 1699 got an automatic review
```
